### 2020-02-03 ChiSubmission/RebuttalCHI2019.txt

Rebuttal
We would like to thank the reviewers for valuable and constructive feedback and appreciate the opportunity to address the main comments 
expressed in them and how we plan to address them in the revised version.

Relevance to CHI:
In regard to the relevance of our work to the CHI community, we agree with the reviewers and will provide a better framing of the work’s 
relevance and potential impact to the CHI community. In particular, we will state that knowledge workers’ stress, and lack of focus or 
awakeness can have a significant impact on their interaction with computers and their work in general. For control room operators, similar to 
car drivers, a lack of focus and awakeness can even lead to severe incidents. We will further state, that various HCI researchers have 
therefore previously examined these and similar variables, in particular, (a) their occurrence, cause and effect in various context, including 
knowledge work, e.g. [50,51], (b) how they can be measured, e.g. [Reb5], and (c) how they might be modulated/regulated, e.g. [Reb7]. While 
previous work mostly focused on smaller or controlled lab studies, our study extends this work by performing a field study to examine whether 
and how these three variables--stress, focus and awakeness--can be measured over an extended period of time in a knowledge workers’ work 
context. We will further state that our results show that a combination of sensors can be used in the field to automatically measure these 
variables, in particular awakeness and stress with a certain accuracy, and that even less invasive computer interaction sensors can be used for 
measuring these variables in certain contexts. In addition, we will extend the discussion to include the usage of such a measurement in the 
workplace, in particular in control rooms as well as regular knowledge worker’s contexts, to modulate stress and awakeness with the goal to 
increase the wellbeing of the worker and the quality of their work, e.g. through individual approaches such as adapting the lighting in a 
control room, or scheduling breaks, and more team based approaches such as an awareness dashboard of a teams’ stress level.

Related work
Regarding the related work, we agree with the reviewers and will extend the comparison of our work to previous work. As stated above, we will 
provide a better framing of our work and situate it better in related work in terms of the study of these variables, the measuring thereof and 
the approaches to regulate/modulate them. We further thank the reviewers for their pointers to relevant related work and will add them. In 
particular, we will also state that previous approaches have already shown the feasibility of applying machine learning to biometric data to 
measure variables, such as stress, interruptibility, negative emotion or confusion during computer interaction [Reb1, Reb3, Reb4] and that 
these approaches vary in the variables examined, the invasiveness of the sensors and the length and context of the studies. We will state that 
many previous studies focused on shorter time frames (mostly hours, but in one case also up to 5 days [Reb2] or 2 weeks [Reb1]), but that none 
examined an 8 week period in a real-world office environment and these variables.

Data collection and preparation:
We thank the reviewers for their comments on the data collection and will clarify this aspect in our paper. We will state that we decided upon 
our data collection method based on prior work [Reb3,Reb4] and through extensive piloting with users. In the end we decided on the text 
messaging approach since pilot participants preferred these over other means also due to them being accessible and noticeable anywhere in the 
office. Biometric data collection was performed with the state-of-the-art Everion device which captures a broad set of signals (Table 1). We 
are aware that there are biometric signals not being captured by this device, or captured at a higher granularity that desired. However, in an 
effort to maintain this research applicable to real-world environments, we picked this already existing device, even when, as a trade-off, we 
could not capture more descriptive and more intrusive signals as SDNN, SCL, SCR, eye tracking, or brain activity.

Technical implementations:
Low precision and recall values in some of the users in Table 3 are due to the unbalanced nature of the responses of these particular users. It 
is worth noticing how users that have the lowest precision and recall values are those where the data is the most unbalanced (column 
“Distribution”, Table 3). This translates to a scenario where the classifier will likely label the minority instances as part of the majority 
class due to the unbalanced ratio of the data.
As to clarify how the data is labelled, the participants do not label each time window analyzed by the features (e.g., the participant does not 
label the last 3 hours as “awake”). The participants respond to how they are currently feeling, and based in the response, we create a set of 
features using different time windows (Figure 5), finally we analyze which features (time windows and biometric signals) can predict best the 
“awake” value.

Reb1. M. Zuger et al. Analyzing Interruptibility in the Office: A Field Study on the Use of Biometric and Computer Interaction Sensors. CHI18
Reb2. A. Sano et al. Stress Recognition using Wearable Sensors and Mobile Phones. ACII13
Reb3. S. Lallé et al. Predicting Confusion in Information Visualization from Eye Tracking and Interaction Data. IJCAI16
Reb4. P. Panwar et al. Detecting Negative Emotion for Mixed Initiative Visual Analytics. CHI18
Reb5. Hernandez et al. Under pressure: Sensing Stress of Computer Users. CHI14
Reb6. G. Mark et al. Email Duration, Batching and Self-interruption: Patterns of Email Use on Productivity and Stress. CHI16
Reb7. J. Snyder et al. MoodLight: Exploring Personal and Social Implications of Ambient Display of Biosensor Data. CSCW15

### 2020-02-03 UbiCompSubmission/bp_conclusion.tex
Adding a new folder UbiCompSubmission with ACM Large format for Ubicomp

%!TEX root = bioPrediction_main.tex
\section{Conclusions} 
Stress, awakeness and focus at work are highly relevant aspects when
it comes to productivity and well-being at the workplace. In this
paper, we presented the results of a study with 14 professional
knowledge workers in their workplace over an eight week period to
better understand how workres experience stress over time and to
examine the ability of biometrics to predict these
productivity-related aspects. The longitudinal and in situ placement of
the study support and extend previous work. Based on
twice collected daily survey responses, we observed that
although participants sometimees saw preiods of sustained strses,
they would always return to a baseline stress level for them
at some point. We also observed that stress levels seldom spiked,
but when they did rise, the rise in stress tended to last more
than a day. In addition to the survey responses, we 
continually collected
biometrics data with which we were able to create a model that was able to
predict awakeness, stress, and focus.

>>>>

%!TEX root = bioPrediction_main.tex
\section{Conclusions} 
Stress, awakeness and focus at work are highly relevant aspects when
it comes to productivity and well-being at the workplace. In this
paper, we presented the results of a study with 14 professional
knowledge workers in their workplace over an eight week period to
better understand how workers experience stress over time and to
examine the ability of biometrics to predict these
productivity-related aspects. The longitudinal and in situ placement of
the study support and extend previous work. Based on
twice collected daily survey responses, we observed that
although participants sometimes saw periods of sustained stress,
they would always return to a baseline stress level for them
at some point. We also observed that stress levels seldom spiked,
but when they did rise, the rise in stress tended to last more
than a day. In addition to the survey responses, we 
continually collected
biometrics data with which we were able to create a model that was able to
predict awakeness, stress, and focus.

### 2020-02-03 UbiCompSubmission/bp_relatedWork.tex
Adding a new folder UbiCompSubmission with ACM Large format for Ubicomp

%!TEX root = bioPrediction_main.tex
\section{Related Work}
We review two categories of related work for this study. First, we review work related to the productivity-related indicators that we considered in our study, including how they are detected and prevented in several risk-inducing scenarios. Second, we review work related to biometrics, including how they have been studied in the office environment or applied to assess cognitive states and processes.

@@

Previous approaches~\cite{zuger18,Lalle16,Panwar18} have shown the feasibility of applying machine learning 
to biometric data to measure variables such as stress, interruptibility, 
negative emotion, or confusion during computer 
interaction. These approaches vary in the 
variables examined, the invasiveness of the sensors, and the length and 
context of the studies. Many previous studies focused on 
shorter time frames (mostly hours, but in one case also up to five 
days~\cite{sano2013stress} 
or two weeks~\cite{zuger18}). To the best of the authors' knowledge, no 
other study has examined such an extensive period (eight weeks) in a real-
world office environment and the inspected variables (stress, awakeness, and focus).

>>>>

%!TEX root = bioPrediction_main.tex
\section{Related Work}

The study and prediction approach on which we report in this paper is related
to previous studies of productivity-related indicators and studies
using biometerics to predict these indicators. We consider
related work in each of these categories in turn.

@@

Our study focuses on the productivity-related indicators of stress, focus and awakeness.

%% Previous approaches~\cite{zuger18,Lalle16,Panwar18} have shown the feasibility of applying machine learning 
%% to biometric data to measure variables such as stress, interruptibility, 
%% negative emotion, or confusion during computer 
%% interaction. These approaches vary in the 
%% variables examined, the invasiveness of the sensors, and the length and 
%% context of the studies. Many previous studies focused on 
%% shorter time frames (mostly hours, but in one case also up to five 
%% days~\cite{sano2013stress} 
%% or two weeks~\cite{zuger18}). To the best of the authors' knowledge, no 
%% other study has examined such an extensive period (eight weeks) in a real-
%% world office environment and the inspected variables (stress, awakeness, and focus).

### 2020-02-03 UbiCompSubmission/bp_results_1.tex
Adding a new folder UbiCompSubmission with ACM Large format for Ubicomp

\subsection{Explaining Stress Fluctuations}
We created a linear mixed model with participants as random effects
to attempt to explain some of the
stress that our participants were experiencing. We experimented with day of the week and proximity to
beginning or end of month as possible explanatory variables. We also calculated
features from the gathered computer interaction data where available - namely total time spent actively on the computer,
percentage of time idle, and amount of time spent on non work-related
web browsing. However,
ultimately we were unable to find any variable which had
significant (p < 0.05) explanatory power with respect to our participants perceived stress
levels. This points to the need for additional instrumentation if we
are to succesfully understand and predict stress in the workplace.

>>>>

\subsection{Explaining Stress Fluctuations}
In an attempt to explain some of the stress that our participants were experiencing, we created a linear mixed model with the self-reported daily stress levels as dependent variable and the participants as random effects. We experimented with day of the week and proximity to beginning or end of month as possible explanatory variables. We also calculated daily features from the gathered computer interaction data where available---namely total time spent actively on the computer, percentage of time idle, and amount of time spent on non work-related web browsing per day. Ultimately, the analysis showed that none of the variables that we examine had a significant explanatory power with respect to our participants perceived stress levels. This indicates that aspects such as the time spent on a computer or the non work-related web browsing do not have a significant impact on the stress level. Furthermore, this points to the need for additional instrumentation and data collection, if we are to succesfully understand and predict stress in the workplace.

### 2020-02-03 UbiCompSubmission/bp_studyMethod.tex
Adding a new folder UbiCompSubmission with ACM Large format for Ubicomp

\subsubsection{Computer Interaction Data}
To gain a better understanding of our participants day-to-day work activities, we installed an open source computer interaction monitor (reference omitted for doubleblin.). The monitor ran in the background on participants' computers and tracked the active windows, as well as the keyboard and mouse activity. Four of our participants opted out of this part of the study for privacy reasons (S6, S10, S12, and S13). Therefore, we only included the computer interaction data in our analysis for our more coarse-grained explanatory model but not in our analysis on the predictive model for stress in the moment.

>>>>

\subsubsection{Computer Interaction Data}
To gain a better understanding of our participants day-to-day work activities, we installed an open source computer interaction monitor (reference omitted for doubleblind). The monitor ran in the background on participants' computers and tracked the active windows, as well as the keyboard and mouse activity. Four of our participants opted out of this part of the study for privacy reasons (S6, S10, S12, and S13). Therefore, we only included the computer interaction data in our analysis for our more coarse-grained explanatory model but not in our analysis on the predictive model for stress in the moment.
